
Introduction to ggplot2
=======================

```{r}
library(ggplot2)
library(reshape2)
```

Basic plotting
--------------

`qplot()` is ggplot2's equivalent of `plot()`

```{r fig.width=8, fig.height=5}
qplot(x=iris$Petal.Width, y=iris$Petal.Length,
      colour=iris$Species, geom='point')
```

Produces the same results as

```{r fig.width=8, fig.height=5, fig.keep='none'}
ggplot(iris, aes(x=Petal.Width, y=Petal.Length, colour=Species)) + geom_point()
```

This command consists of three main components:

* The data, `iris`, which is `dataframe` (or a sub-class, like `data.table`).
* A set of mappings from columns of the data set, `aes(...)`.
* A geometry, `geom_histogram()`.

What's going on?
----------------

ggplot2's language is based on the "Grammar of graphics" (Wilkinson 1999).
Every plot consists of a background, and one or more of each of the following:

* **Layers** - sets of data (from a `dataframe` or `data.table`) with mappings, scales, geometries and statistics applied. Data needs to be in 'long; format - one
* **Aesthetic mappings** - conversion from a data variable (Sepal.Width, ...) to a visual variable (x, y, colour, size, ...).
* **Scales** - functions, limits, and divisions of the visual variables (logarithmic x-axis, colours gradients, y-axis division lines, ...).
* **Geoms** - methods of plotting the data (points, lines, boxplots, ...).
* **Statistics** - conversions between the raw data and some visual summary (**identity**, binning, quantiles, smoothers, ...)

Most geoms are actually a shortcut for a set of default scales, lower-level geometries, and statistics. Most statistics have default geoms. Each of these automatically set up a layer, so you don't have to specify it manually.

There are also a few other components that pop up less frequently, including coordinate systems (e.g. polar plots), positions (fine tuning of object positioning), facets (small-multiple plotting), and themes (that control various graphical aspects of the whole plot).

There is a full list of available geoms, stats, etc. at http://docs.ggplot2.org/current/ .
Also, all ggplot functions have good built-in docs, which list the default stats for geoms, etc.


### Example

We can create histograms similarly to the plots up the top:

```{r fig.width=8, fig.height=5}
ggplot(data=iris, aes(x=Petal.Width, fill=Species)) + geom_histogram(binwidth=0.2)
```

What's happening:

* We're using `iris` as our main data source.
* We're using the histogram geom.
  * geom_histogram uses `stat_bin()` as it's default statistic.
    * `stat_bin` puts up a warning about binwidth if you don't include the `binwidth` parameter. geom_histogram helpfully passes it along.
* We're mapping the x axis to `Petal.Width` in the `iris` dataframe.
  * geom_histogram handles the mapping from the y axis to the counts for each bin from `stat_bin()`.
* We're mapping the colour of the bars (actually geom_rectangles in disguise) to the Species in `iris`.
  * geom_histogram uses `position_stack()` by default, so the colours are stacked.
* ggplot uses linear scales for all continuous/ordinal variables by default. For categorical variables (e.g. Species) it divides the space (e.g. colour==hue) up into equal divisions.


Here is a fully specified version of the plot above:

```{r fig.width=8, fig.height=5, fig.keep='none'}
ggplot() +
  layer(data=iris,
       mapping=aes(x=Petal.Width,
                   fill=Species
                   ),
       stat='bin',
       stat_params=list(binwidth=0.2),
       geom='histogram',
       geom_params=list(),
       position='identity'
       ) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_fill_hue() +
  coord_cartesian()
```


### Modifying plots

We can change any aspect of this plot, by adding the relevant scales/positions/statistics etc.:

```{r fig.width=8, fig.height=5}
p <- ggplot(iris, aes(x=Petal.Width, fill=Species)) +
  geom_histogram(binwidth=0.2, position=position_dodge()) +
  scale_fill_manual(values=c('red', 'black', 'green'))
p
```

In the above example, we stored the plot in the variable `p`. We can create variations on the plot by adding more ggplot functions:

```{r fig.width=8, fig.height=5}
p + scale_y_continuous(name='Number of samples') + scale_x_continuous(name="Petal width") + coord_polar()
```

If you add a scale that already exists, the old one gets overwritten (with a warning). So:

```{r fig.width=8, fig.height=5}
p + scale_fill_grey()
```


Facetting
---------

Faceting is a really handy way to create small-multiples sub-plots. There are two options: `facet_grid()`, and `facet_wrap()`.

```{r fig.width=8, fig.height=5}
# help(CO2) # for a description
head(CO2, 3)

ggplot(CO2, aes(y=uptake, x=conc)) + geom_line()
```

Doesn't work because ggplot treats all data as a single set. We can tell it to treat separate groups as subsets:

```{r fig.width=8, fig.height=5}
p <- ggplot(CO2, aes(y=uptake, x=conc, group=Plant)) + geom_line()
p
```

`aes(colour=...)` implies group too, but isn't that useful here:

```{r fig.width=8, fig.height=5}
ggplot(CO2, aes(y=uptake, x=conc, colour=Plant)) + geom_line()
```

Instead, we can create a set of small-multiples over the type and treatment factors:

```{r fig.width=8, fig.height=5}
p + facet_grid(Type ~ Treatment, scale='free_y')
```


Plotting non-dataframe data:
----------------------------

The `volcano` dataset is a matrix of heights:

```{r}
str(volcano)
```

Using `melt()` from the `reshape2` library, we can convert it to a long dataframe:

```{r fig.width=8, fig.height=5}
volc_m <- melt(volcano, varnames=c('lat', 'long'), value.name='height')

str(volc_m)

p_volcano <- ggplot(volc_m, aes(x=lat, y=long, fill=height)) + geom_tile()
p_volcano
```

Poor colours, we can improve those:

```{r fig.width=8, fig.height=5}
p_volcano + scale_fill_gradientn(colours=terrain.colors(10))
```

We could also add some contour lines:

```{r fig.width=8, fig.height=5}
p_volcano + scale_fill_gradientn(colours=terrain.colors(10)) +
  geom_contour()
```

Doesn't work, because we haven't provided a mapping for z - the apparent "height" of the tiles was defined by fill colour, not z. So we add another mapping:

```{r fig.width=8, fig.height=5}
p_volcano + scale_fill_gradientn(colours=terrain.colors(10)) +
  geom_contour(mapping=aes(z=height), colour='black')
```

This mapping only applies to the stat_contour. We could also have added it to the original `ggplot(aes(...))`, and it would apply to all layers/geoms (which wouldn't change anything in this case, because `geom_tile` doesn't have a z variable).

#### Converting from wide-format dataframes:

From wide data: USJudgeRatings:

```{r}
head(USJudgeRatings, 3)

# convert row names to a data frame variable
USJudgeRatings$name <- rownames(USJudgeRatings)
head(USJudgeRatings, 3)

# Just use a subset for a simpler plot
USJR_melt <- melt(USJudgeRatings[,])
head(USJR_melt, 3)
```

Now we can plot it!

```{r fig.width=10, fig.height=7}
ggplot(USJR_melt, aes(x=variable, y=value, colour=value)) +
  geom_point() +
  geom_line(aes(x=as.integer(variable))) +
  scale_colour_gradientn(colours=heat.colors(5)) +
  facet_wrap('name', ncol=5) +
  theme_bw() +
  theme(axis.text.x=element_text(angle=-90))
```

`theme()` lets you pass various options to control global plotting variables. `help(theme)` has a big list of them.

Positions
----------

```{r fig.width=8, fig.height=5}
p_mpg <- ggplot(mpg, aes(x=hwy, y=cty, colour=displ))
p_mpg + geom_point()
```

Some overlap here, so we can add some jitter:

```{r fig.width=8, fig.height=5}
p_mpg + geom_point(position=position_jitter())
```

Could also plot the density:

```{r fig.width=8, fig.height=5}
p_mpg + geom_point(position=position_jitter()) +
  geom_density2d(colour='black', alpha=0.3)
```


#### Smoothing

It might be more helpful to just plot points and an average here:

```{r fig.width=8, fig.height=5}
# leave group out, so that the smoother uses all data:
p <- ggplot(CO2, aes(y=uptake, x=conc)) + geom_point() +
  facet_grid(Type ~ Treatment)
p + stat_smooth() # geom_smooth does the same thing there: same defaults
```

That smoother is the default, which is loess. Loess can be a little too detailed though, and it's possible to pass any smoother, for instance, `lm`:

```{r fig.width=8, fig.height=5}
p <- ggplot(CO2, aes(y=uptake, x=conc)) + geom_point() +
  facet_grid(Type ~ Treatment)
p + stat_smooth(method='lm')
```

Probably not enough detail. How about just the mean at each concentration, plus a standard deviation?

```{r fig.width=8, fig.height=5}
# from http://stackoverflow.com/questions/12033319
p + stat_summary(fun.y = 'mean', colour = 'blue', geom = 'line') +
    stat_summary(fun.data = 'mean_cl_normal', geom = 'ribbon', alpha = 0.25)
```

Mapping
-------

Mapping can be done over the top of google maps with the `ggmap` package.

```{r fig.width=8, fig.height=5}
# based on http://stackoverflow.com/questions/10302195
library(ggmap)

quakes_map <- get_map(location=c(lon=mean(quakes$lon), lat=mean(quakes$lat)), zoom = 5, maptype = 'satellite')

ggmap(quakes_map) +
  geom_point(data=quakes, mapping=aes(x=long, y=lat, size=mag, colour=depth)) +
  scale_colour_gradientn(colours=heat.colors(5))

```

You can add`coord_map()` to use different map projections, but it doesn't play well with ggmap by default.

